---
title: 'Soil properties mediated by altitude, not geography or climatic distance, explain the distribution of a tropical endemic herb'
author: "Jacob Moutouama and Orou Gaoue"
output: 
  rmdformats::readthedown:
    code_folding: show
    self_contained: true
    number_sections: true
    thumbnails: true
    lightbox: true
    gallery: true
    keep_md: true
    highlight: tango
    df_print: kable 
    toc_depth: 3
    fig_width: 8
    fig_height: 8
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(external=TRUE,echo=F,warning=FALSE,fig.pos='H', global.device = TRUE)
a4width<- 8.3
a4height<- 11.7
options(knitr.table.format = "latex")
```

This document provides code to reproduce the analyses from:

> Moutouama, J.K. and OG Gaoue. 2021. "Soil properties mediated by altitude, not geography or climatic distance, explain the distribution of a tropical endemic herb" _Ecology and Evolution_


```{r eval=TRUE,echo=T}
rm(list = ls())
source("/Users/jmoutouama/Dropbox/PhD Project/Biotic_abiotic_Thunbergia/load_pkgs.R")
```

# Soil fertility data to reduce to data to Principal component 

```{r, results="hide"}
Thunbergia.soil<-read.csv("/Users/jmoutouama/Desktop/Thunbergia /Data/Soil.csv", header=T)
str(Thunbergia.soil)
head(Thunbergia.soil)
summary(Thunbergia.soil)
Soil.pca<-PCA(Thunbergia.soil[,-c(1,2)], scale.unit = TRUE, ncp = 5, graph = FALSE) 
```

```{r}
fviz_eig(Soil.pca, addlabels = TRUE, ylim = c(0, 50))
```


```{r}
library(corrplot)
var <- get_pca_var(Soil.pca)
head(var$contrib, 4)
corrplot(var$contrib, is.corr=FALSE) 
```


```{r}
fviz_pca_var(Soil.pca, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")
             )
```

```{r}
ind <- get_pca_ind(Soil.pca)
ind.contribution<-ind$contrib
ind.cordinate<-ind$coord
data.soil<-cbind.data.frame(Thunbergia.soil,ind.cordinate) 
#write.infile(data.soil, "/Users/jmoutouama/Desktop/Thunbergia /Data/data.soil.csv", sep = ";")
```

```{r}
Soil.plot<-fviz_pca_biplot(Soil.pca, 
                # Individuals
                geom.ind = "point",
                fill.ind = Thunbergia.soil$Population, col.ind = "black",
                pointshape = 21, pointsize = 2,
                palette = NULL,
                addEllipses = TRUE,
                # Variables
                alpha.var ="contrib", col.var = "contrib",
                gradient.cols = c("white", "blue", "red"),
                title = "",
                legend.title = list(fill = "Population", color = "Contribution")
                )

Soil.plot
#ggsave("/Users/jmoutouama/Desktop/Thunbergia/Figure/Soil.plot.png", Soil.plot, width = 10,height = 10,dpi = 600)



```

The calculation of the geographic distance was obtained from: 
> Dallas, T., Decker, R. R., & Hastings, A. (2017). Species are not most abundant in the centre of their geographic range or climatic niche. Ecology Letters, 20(12), 1526–1533. https://doi.org/10.1111/ele.12860

# Centroid of geographic range 

```{r eval=T,results="hide"}
library(geosphere)
Thunbergia <- Thunbergia[,3:4] 
## estimate the centroid
centroid(Thunbergia)
#Thunbergia_centroid<-rbind(Thunbergia,c(10.07482,1.692276))
Thunbergia_centroid<-rbind(Thunbergia,c(9.925417,1.801845))

## Distance from the geographic center
result <- data.frame(matrix(nrow = 55, ncol = 2))
colnames(result) <- c("i", "distance")
for (i in 1:55) {
  dis<-distGeo(Thunbergia_centroid[i, ], Thunbergia_centroid[55, ])
  result[i, 1] <- i
  result[i,2]<-dis
}

result
#write.infile(result, "distance.center.csv", sep = ";")

```

The calculation of the climatic distance (here mahanalobis distance) was obtained from: 

> Luis Osorio-Olvera, Lira-Noriega, A., Soberón, J., Peterson, A. T., Falconi, M., Contreras-Díaz, R. G., … Barve, N. (2020). ntbox: an R package with graphical user interface for modeling and evaluating multidimensional ecological niches. Methods in Ecology and Evolution, 1(1), 1–2. doi:10.1111/j.2041-210x.2010.00016.x.
# Model calibration selection using Minimum Volume Ellipsoids (MVEs).

```{r,results="hide"}
set.seed(100)

bioclim<-stack(list.files("/Users/jmoutouama/Dropbox/PhD Project/Biotic_abiotic_Thunbergia/bioclim",
                      pattern = ".tif$",
                      full.names = T))
plot(bioclim)

```

```{r}
# Let's  split the occurrence data into train and test using the type variable; the choice of the data is  done using different random partition methods.

thunb<- read.csv("/Users/jmoutouama/Dropbox/PhD Project/Biotic_abiotic_Thunbergia/Data/occurence.convexhull.csv", stringsAsFactors=FALSE)

# Random sample indexes
train_index <- sample(1:nrow(thunb), 0.8 * nrow(thunb))
test_index <- setdiff(1:nrow(thunb), train_index)


# # Split occs in train and test
thunb_train <- thunb[train_index, ]
thunb_test <- thunb[test_index, ]


#The following code extracts the environmental information for both train and test data

thunb_etrain <- raster::extract(bioclim,thunb_train[,c("longitude", "latitude")],df=TRUE)
thunb_etrain <- thunb_etrain[,-1]
head(thunb_etrain)


thunb_etest <- raster::extract(bioclim,thunb_test[,c("longitude","latitude")], df=TRUE)
thunb_etest <- thunb_etest[,-1]
head(thunb_etest)
```

```{r}
env_varsL <- ntbox::correlation_finder(cor(thunb_etrain),threshold = 0.8,verbose = F)
env_vars <- env_varsL$descriptors
print(env_vars )

# Now we specify the number of variables to fit the ellipsoid models; in the example, we will fit for 3,5, and 6 dimensions
nvarstest <- c(3,5,6)

# This parameter is to specify the proportion of training points that will be used to fit the minimum volume ellipsoid (Van Aelst and Rousseeuw 2009).


# Level
level <- 0.99


# This background data is just to compute the partial ROC test
env_bg <- ntbox::sample_envbg(bioclim,10000)

# For selecting the model we will use an arbitrary value of 6 percent of omission; it is not a rule but accepted omission rates are those bellow 10%. We will ask the function to return the partial ROC value (Peterson, Papes, and Soberon 2008)

omr_criteria <- 0.06
proc <- TRUE

# Now we just need to use the function ellipsoid_selection to run the model calibration and selection protocol

e_selct <- ntbox::ellipsoid_selection(env_train = thunb_etrain,
                                      env_test = thunb_etest,
                                      env_vars = env_vars,
                                      level = level,
                                      nvarstest = nvarstest,
                                      env_bg = env_bg,
                                      omr_criteria= omr_criteria,
                                      proc = proc)


# Let’s see the first 20 rows of the results
head(e_selct,20)


# With the following lines of code, I am going to display the model in the first row of the table
# Best ellipsoid model for "omr_criteria" 
bestvarcomb <- stringr::str_split(e_selct$fitted_vars,",")[[1]]





# Ellipsoid model (environmental space)
best_mod <- ntbox::cov_center(thunb_etrain[,bestvarcomb],
                              mve = T,
                              level = 0.99,
                              vars = 1:length(bestvarcomb))

class(best_mod)


# Projection model in geographic space

mProj <- ntbox::ellipsoidfit(bioclim[[bestvarcomb]],
                             centroid = best_mod$centroid,
                             covar = best_mod$covariance,
                             level = 0.99,size = 3)
dev.new(
  title = "Chord",
  width = 10,
  height = 5,
  noRStudioGD = TRUE
)

raster::plot(mProj$suitRaster)
points(thunb[,c("longitude","latitude")],pch=20,cex=0.5)


thunb_plot<- Thunbergia
thunb_plot<- raster::extract(bioclim,thunb_plot[,c("longitude","latitude")], df=TRUE)
thunb_plot <- thunb_plot[,-1]

mhd <- stats::mahalanobis(thunb_plot[,bestvarcomb],center = best_mod$centroid,cov = best_mod$covariance)

distance<-data.frame(mhd)

mahanalobis<-cbind(Thunbergia,distance)

plot(Density.per.plot~mhd,data=mahanalobis)
mod<-lm(Density.per.plot~mhd,data=mahanalobis)
summary(mod)
# write.infile(mahanalobis, "/Users/jmoutouama/Dropbox/PhD Project/Biotic_abiotic_Thunbergia/Data/distance.center.csv", sep = ";")


```

# PiecewiseSEM 
```{r,results="hide"}
##  Load data 
Thunbergia.sem<-read.csv("/Users/jmoutouama/Dropbox/PhD Project/Biotic_abiotic_Thunbergia/Data/Thunbergia.csv", header=T)
str(Thunbergia.sem)
names(Thunbergia.sem)
dim(Thunbergia.sem)
Thunbergia.sem$Plot<-factor(Thunbergia.sem$Plot)
for(i in c("LMA","skewness")) Thunbergia.sem[,i] <- as.numeric(Thunbergia.sem[,i])
str(Thunbergia.sem)
summary(Thunbergia.sem)
zThunbergia.sem <- scale(Thunbergia.sem[,4:14])
density<-log(Thunbergia.sem$Density)
zThunbergia.sem<-data.frame(Thunbergia.sem[,c(1,2,15,16)],zThunbergia.sem,density)
summary(zThunbergia.sem)
names(zThunbergia.sem)[which(names(zThunbergia.sem) == "mhd.1")] <- "mhd"
```


## Model including distance from geographic center and altitude 
```{r}
library(lme4)
library(piecewiseSEM)
library(nlme)
corsem_final <- piecewiseSEM::psem(
  lme(PCA1 ~ Distance.center+Altitude,random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  lme(PCA2 ~ Distance.center+Altitude,random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  lme(LMA ~ PCA1+ Distance.center+ PCA2 + SRU +PAR +Altitude+Soil.moisture, random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  lme(Chlorophyle ~ PCA1+ Distance.center+ PCA2 + SRU +PAR +Altitude+Soil.moisture, random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  lme(PAR ~ Distance.center+Altitude,random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  lme(Soil.moisture ~ Distance.center+Altitude+SRU,random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  lme(SRU ~ PAR+Distance.center+Altitude+PCA1+PCA2+Chlorophyle+LMA+Soil.moisture,random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  lme(basal.diameter ~ PAR+Distance.center+Altitude+PCA1+ PCA2+SRU+Chlorophyle+LMA+density+Soil.moisture,random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  lme(density ~ PAR+Distance.center+Altitude+PCA1+ PCA2+SRU+Chlorophyle+LMA+basal.diameter+Soil.moisture+skewness,random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  lme(skewness ~ PAR+Distance.center+Altitude+PCA1+ PCA2+SRU+Chlorophyle+LMA+basal.diameter+Soil.moisture,random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  Distance.center %~~% Altitude,
  Soil.moisture %~~%  PCA1,
  Soil.moisture %~~%  PCA2
)

```

```{r}
d<-summary(corsem_final, .progressBar = F)
d$coefficients
```

## Model including distance from  climatic niche center and altitude 

```{r}
corsem_final_ellipsoid <- psem(
  lme(PCA1 ~ mhd+Altitude,random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  lme(PCA2 ~ mhd+Altitude,random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  lme(LMA ~ PCA1+ mhd+ PCA2 + SRU +PAR +Altitude+Soil.moisture, random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  lme(Chlorophyle ~ PCA1+ mhd+ PCA2 + SRU +PAR +Altitude+Soil.moisture, random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  lme(PAR ~ mhd+Altitude,random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  lme(Soil.moisture ~ mhd+Altitude,random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  lme(SRU ~ PAR+mhd+Altitude+PCA1+PCA2+Chlorophyle+LMA+Soil.moisture,random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  lme(basal.diameter ~ PAR+mhd+Altitude+PCA1+ PCA2+SRU+Chlorophyle+LMA+density+Soil.moisture,random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  lme(density ~ PAR+mhd+Altitude+PCA1+ PCA2+SRU+Chlorophyle+LMA+basal.diameter+Soil.moisture,random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  lme(skewness ~ PAR+mhd+Altitude+PCA1+ PCA2+SRU+Chlorophyle+LMA+basal.diameter+Soil.moisture,random = ~1|Population, na.action = na.omit, data = zThunbergia.sem),
  Distance.center %~~% Altitude,
  Soil.moisture %~~% PCA1,
  Soil.moisture  %~~%  PAR
)

```

```{r}
b<-summary(corsem_final_ellipsoid, .progressBar = F)
b$coefficients
```

